Summer 2025 ToDo

### 2025-07-23 Relations/Paper/Well-foundedness.tex
WF edits

@@

For example, the implication from ``every non-empty subset has a minimal element'' to
``every element is accessible"  is a classical proof by contradiction.
(Assuming that there exists an inaccessible element, the hypothesis implies that
there must be a minimal such element, which immediately yields a contradiction:
every minimal element is accessible.)

@@

One could also consider proving well-foundedness ``up to $\lnot\lnot$",
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
On the other hand, the implication from the sequential to inductive definition
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {\accCor} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  Assuming \accCor, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

As this discussion illustrates, the network of relationships between the
constructive notions of well-foundedness is quite rich.  The purpose of
this section is to present the basic picture of these implications,
and elaborate on some of them.  Our key observations are summarized
in the following list.
\begin{enumerate}
  \item The implication from the inductive to the sequential definition
  requires the underlying relation to be decidable. This assumption is no longer
  necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires Markov's Principle
  (and decidability of $R$) to yield the same consequence.
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent,
  and imply WFacc- via WFseq-, requiring {\accCor} assumption in the final inference.
  \item At the same time, if the relation is finitely branching, then WFminDNE-
  implies WFacc- directly.
  \item Finite branchingness together with (weak) decidability of minimality
  implies \accCor.
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  \item Assuming \accCor, isWFacc- can also be derived from a weaker condition, isWFminCor(-).
  (Just like isWFminDNE is a natural weakening of isWFmin sufficient to recover isWFacc
  when {\accDNE} is available.)
  It is implied by isWFseq-, but does not appear to imply anything on its own.

>>>>

\apcomment{Well-foundedness or wellfoundedness?}

@@

For example, the implication from ``every non-empty subset has a minimal element'' to
``every element is accessible''  is a classical proof by contradiction.
(Assuming that there exists an inaccessible element, the hypothesis implies that
there must be a minimal such element, which immediately yields a contradiction:
every minimal element is accessible.)

@@

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the
implication from sequential to inductive wellfoundedness
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {\accCor} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  (In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.)
Assuming \accCor, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

As this discussion illustrates, the network of relationships between the
constructive notions of well-foundedness is quite rich.  The purpose of
this section is to present the basic picture of these implications,
and elaborate on some of them.  Our key observations are summarized
in the following list.
\begin{enumerate}
  \item The implication from the inductive to the sequential definition
  requires the underlying relation to be decidable. This assumption is no longer
  necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires Markov's Principle
  (and decidability of $R$) to yield the same consequence.
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent,
  and imply WFacc- via WFseq-, requiring {\accCor} assumption in the final inference.
  \item At the same time, if the relation is finitely branching, then WFminDNE-
  implies WFacc- directly.
  \item Finite branchingness together with (weak) decidability of minimality
  implies \accCor.
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  \item Assuming \accCor, isWFacc- can also be derived from a weaker condition, isWFminCor(-).
  (Just like isWFminDNE is a natural weakening of isWFmin that is sufficient to recover isWFacc
  in the presence of {\accDNE}.)
  It is implied by isWFseq-, but does not appear to imply anything on its own.
